---
editor_options: 
  markdown: 
    wrap: 72
---

## Exercise 1

Begin with describing and fitting a full model in which the intercepts
and slopes of the extinction times versus numbers of pairs may be
different in all four combinations of size and migratory status.

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

```{r, echo=FALSE}
# Load the dataset
bird_data <- read.csv("data.csv")

# Fit the full model
full_model <- lm(Time ~ Pairs * Size * Status, data = bird_data)

# Display the summary of the model
summary(full_model)
```

```{r, echo=FALSE}
# Add the fitted values to the dataset
bird_data$fitted_values <- fitted(full_model)

# Plot the data with fitted lines
ggplot(bird_data, aes(x = Pairs, y = Time, color = interaction(Size, Status))) +
  geom_point() +
  geom_line(aes(y = fitted_values)) +
  labs(title = "Extinction Time vs. Number of Pairs",
       x = "Number of Pairs",
       y = "Extinction Time (Years)",
       color = "Size & Status") +
  theme_minimal()
```

## Exercise 2

Examine the data set for possible transformations and outliers: Make a
residual plot from the fit of the model. What can you tell based on the
residual plot? (i.e. fit of the model/ possible transformations needed +
outliers)?

```{r, echo=FALSE}
# Calculate residuals
bird_data$residuals <- residuals(full_model)

# Create a residual plot
ggplot(bird_data, aes(x = fitted_values, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Plot",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```

When examining the residuals, we look for:

1.  Homoscedasticity: The residuals should have constant variance across
    the range of fitted values. If the spread of residuals seems to
    change across the fitted values, a transformation may be needed.

The variance of the residuals becomes bigger and bigger as the value
being fitted (number of pairs + extinction time) goes up.

2.  Independence: There should be no patterns or trends in the
    residuals. If you notice any patterns or trends, it might indicate
    that a variable is missing from the model, or a transformation is
    needed.

There appears to be a slight downward slope pattern.

3.  Outliers: Look for points that stand out from the rest, as they may
    be outliers. Investigate these points further to determine if they
    are errors or if they represent genuine observations. You may
    consider removing outliers if they are affecting the model's
    performance.

Biggest outlier:

-   Species: Raven
-   Time: 58.82
-   Pairs: 2.35
-   Size: L
-   Status: R

4.  Normality: Ideally, the residuals should be approximately normally
    distributed. If the residuals exhibit a non-normal distribution, you
    may need to consider a transformation of the dependent variable or
    use a different modeling approach.

```{r, echo=FALSE}
# Calculate standardized residuals
bird_data$std_residuals <- rstandard(full_model)

# Create Q-Q plot
ggplot(bird_data, aes(sample = std_residuals)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of Standardized Residuals",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

The residuals appear to be normally distributed except for the lower and
upper ends, where they become more skewed, this suggests that a data
transformation may be necessarry.

## Exercise 3

```{r, echo=FALSE}
# Fit the full model with log transformation
log_model <- lm(log(Time) ~ Pairs * Size * Status, data = bird_data)

# Fit the full model with square root transformation
sqrt_model <- lm(sqrt(Time) ~ Pairs * Size * Status, data = bird_data)

# Fit the full model with reciprocal transformation
reciprocal_model <- lm(I(1 / Time) ~ Pairs * Size * Status, data = bird_data)

# Calculate residuals
bird_data$log_residuals <- residuals(log_model)
bird_data$sqrt_residuals <- residuals(sqrt_model)
bird_data$reciprocal_residuals <- residuals(reciprocal_model)
```

```{r, echo=FALSE}
# Log transformation residual plot
ggplot(bird_data, aes(x = fitted(log_model), y = log_residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Plot (Log Transformation)",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```

```{r, echo=FALSE}
# Square root transformation residual plot
ggplot(bird_data, aes(x = fitted(sqrt_model), y = sqrt_residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Plot (Square Root Transformation)",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```

```{r, echo=FALSE}
# Log transformation residual plot
ggplot(bird_data, aes(x = fitted(reciprocal_model), y = reciprocal_residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Plot (Reciprocal Transformation)",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```

Immediately, we can tell that the square transformation does not improve
the model. To compare the log and reciprocal transformation, we will use
Q-Q plots again.

```{r, echo=FALSE}
# Calculate standardized residuals
bird_data$std_log_residuals <- rstandard(log_model)
bird_data$std_reciprocal_residuals <- rstandard(reciprocal_model)
```

```{r, echo=FALSE}
# Create Q-Q plot for the Log Model
ggplot(bird_data, aes(sample = std_log_residuals)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of Standardized Residuals for the Log Transformed Model",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

```{r, echo=FALSE}
# Create Q-Q plot for the Reciprocal Model
ggplot(bird_data, aes(sample = std_reciprocal_residuals)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of Standardized Residuals for the Reciprocal Transformed Model",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

Based on the Q-Q plots, the log transformation seems to be the best fit
for the majority of the fitted data points.

## Exercise 4

Outliers can have a significant impact on the results of a statistical
analysis, including regression models. There is one outlier with nearly
60 years before extinction and possibly other outliers at the other end
of the scale. Here's how these outliers can affect the results:

1.  Influence on model parameters: Outliers can skew the estimates of
    the regression coefficients, resulting in biased estimates of the
    relationship between the independent and dependent variables. This
    can lead to incorrect conclusions about the strength and direction
    of the associations.

2.  Reduced model accuracy: Outliers can increase the overall error in
    the model, reducing its accuracy and predictive power. This can make
    it difficult to generalize the model to new data or other similar
    situations.

3.  Distortion of the residual distribution: Outliers can cause the
    residuals to deviate from normality, violating the assumptions of
    linear regression models. This can affect the reliability of
    hypothesis tests and confidence intervals derived from the model.

Despite these potential issues, there might be valid reasons to keep the
outliers in the dataset:

1.  Real-world representation: If the outliers are genuine observations
    and not a result of data entry errors or other mistakes, they may
    represent important real-world phenomena that should be considered
    in the analysis. Removing them could lead to an incomplete or biased
    understanding of the underlying processes.

2.  Insight into extreme cases: Outliers can provide valuable insights
    into the behavior of the system under extreme conditions. By keeping
    them in the dataset, you can learn more about how the system
    responds when pushed to its limits.

3.  Robustness of the model: Including outliers in the analysis can help
    assess the robustness of the model. If the model can account for the
    presence of outliers and still provide meaningful results, it
    demonstrates the model's ability to handle a wider range of
    situations.

In conclusion, it is essential to carefully assess the potential impact
of outliers on the analysis and consider their potential benefits before
deciding whether to keep them in the dataset. In some cases, it might be
necessary to use robust statistical methods or other techniques to
account for the presence of outliers and minimize their influence on the
results.

## Exercise 5

```{r, echo=FALSE}
# Create a new variable for log-transformed time
bird_data$log_time <- log(bird_data$Time)

# Create scatterplots for each combination of size and migratory status
ggplot(bird_data, aes(x = Pairs, y = log_time, color = Status)) +
  geom_point() +
  facet_wrap(~ Size) +
  labs(title = "Scatterplots of Log(Time) vs. Pairs",
       x = "Pairs",
       y = "Log(Time)") +
  theme_minimal()
```

The relationship between log(Time) and Pairs does appear to be a
positive linear relationship, thus indicating that a transformation of
Pairs is not needed.

The relationship also appears to be more linear than with
log-transformed Pairs, as shown below, where a slight curve pattern
emerges.

```{r, echo=FALSE}
# Create a new variable for log-transformed time
bird_data$log_pairs <- log(bird_data$Pairs)

# Create scatterplots for each combination of size and migratory status
ggplot(bird_data, aes(x = log_pairs, y = log_time, color = Status)) +
  geom_point() +
  facet_wrap(~ Size) +
  labs(title = "Scatterplots of Log(Time) vs. Log(Pairs)",
       x = "Log(Pairs)",
       y = "Log(Time)") +
  theme_minimal()
```

## Exercise 6

### 6.1

```{r, echo=FALSE}
summary(log_model)
```

After running this code, you will see the coefficients for each term in
the model, including the interactions. To determine whether the slopes
are equal for all combinations of "size" and "migratory status," examine
the coefficients for the interaction terms Pairs:Size, Pairs:Status, and
Pairs:Size:Status. If the interaction terms are not statistically
significant (i.e., their p-values are larger than a chosen significance
level, typically 0.05), it suggests that the slopes are not
significantly different between the four combinations of "size" and
"migratory status."

Pairs:SizeS: The coefficient is -0.1089 with a p-value of 0.5266, which
is not statistically significant at a 0.05 significance level.

Pairs:StatusR: The coefficient is -0.1546 with a p-value of 0.2790,
which is not statistically significant at a 0.05 significance level.

Pairs:SizeS:StatusR: The coefficient is 0.2717 with a p-value of 0.1629,
which is not statistically significant at a 0.05 significance level.

Since none of the interaction terms are statistically significant, there
is not enough evidence to conclude that the slopes for all four
combinations of "size" and "migratory status" are different. This
suggests that the relationship between "Pairs" and log("Time") may not
be significantly different among the four groups based on "size" and
"migratory status". However, it is essential to note that a lack of
statistical significance does not necessarily mean the slopes are equal;
it indicates that there isn't enough evidence to reject the null
hypothesis that the slopes are equal.

### 6.2

Nested models are a series of models where each model is a subset of the
previous one. This approach helps in assessing the contribution of
variables and their interactions to the overall model fit. To create
nested models, start with the simplest model and gradually add variables
and interaction terms to evaluate their contributions.

```{r, echo=FALSE}
# Fit the nested models
model1 <- lm(log(Time) ~ Pairs, data = bird_data)
model2 <- lm(log(Time) ~ Pairs + Size, data = bird_data)
model3 <- lm(log(Time) ~ Pairs + Size + Status, data = bird_data)
model4 <- lm(log(Time) ~ Pairs * Size + Status, data = bird_data)
model5 <- lm(log(Time) ~ Pairs * Size * Status, data = bird_data)

# Perform an ANOVA to compare the nested models
anova(model1, model2, model3, model4, model5)
```

When we move from Model 1 to Model 2 (adding the Size variable), the
p-value is 0.000181 (significant at the 0.05 level). This indicates that
adding the Size variable significantly improves the model fit.

When we move from Model 2 to Model 3 (adding the Status variable), the
p-value is 0.007825 (significant at the 0.05 level). This suggests that
adding the Status variable significantly improves the model fit.

When we move from Model 3 to Model 4 (adding the Pairs \* Size
interaction term), the p-value is 0.202995 (not significant at the 0.05
level). This suggests that adding the Pairs \* Size interaction term
does not significantly improve the model fit.

When we move from Model 4 to Model 5 (adding the Pairs \* Size \* Status
interaction term), the p-value is 0.493550 (not significant at the 0.05
level). This indicates that adding the Pairs \* Size \* Status
interaction term does not significantly improve the model fit.

Based on the ANOVA results, it seems that Model 3 (Time \~ Pairs +
Size + Status) is the most appropriate model for this dataset, as adding
interaction terms does not significantly improve the model fit.

## Exercise 7

Based on the findings from previous items, we can create a reduced model
using just the Pairs, Size, and Status variables, without interaction
terms.

```{r, echo=FALSE}
# Reduced Model
reduced_model <- lm(log(Time) ~ Pairs + Size + Status, data = bird_data)

# Summary of the reduced model
summary(reduced_model)
```

```{r, echo=FALSE}
# Add the fitted values to the dataset
bird_data$reduced_fitted_values <- fitted(reduced_model)

# Plot the data with fitted lines
ggplot(bird_data, aes(x = Pairs, y = Time, color = interaction(Size, Status))) +
  geom_point() +
  geom_line(aes(y = reduced_fitted_values)) +
  labs(title = "Extinction Time vs. Number of Pairs",
       x = "Number of Pairs",
       y = "Natural logarithm of Extinction Time (Years)",
       color = "Size & Status") +
  theme_minimal()
```

## Exercise 8

After analyzing the data and accounting for the number of nesting pairs,
we can conclude that the number of nesting pairs has a significant
impact on the time it takes for a species to become extinct. Larger
numbers of nesting pairs tend to result in longer times before
extinction.

However, when considering the size and migratory status of species,
their individual effects on extinction time are less clear. We could not
establish a strong relationship between these factors and the time to
extinction, indicating that they may not be as influential as the number
of nesting pairs.

It is also worth noting that there were a few outliers with unusually
large extinction times compared to other species with similar
explanatory variable values. These outliers should be further
investigated to understand the underlying factors contributing to their
atypical extinction times.

Biggest outlier:

-   Species: Raven
-   Time: 58.82
-   Pairs: 2.35
-   Size: L
-   Status: R

Based on the reduced model, the "theoretical" regression formula for the
logarithm of extinction time can be constructed using the coefficients
from the summary. The formula would look like this:

$$ln(Time) = 0.43056 + 0.26509 * Pairs - 0.65237 * SizeS + 0.50406 *
StatusR$$

Here,

ln(Time) represents the natural logarithm of the extinction time
(note that log()) in R defaults to ln;

Pairs
is the number of nesting pairs (as a continuous variable);

SizeS is a binary variable indicating the bird size, with 1 for
small-sized birds and 0 for large-sized birds;

StatusR is a binary variable indicating the migratory status, with 1 for
resident birds and 0 for migratory birds.

The coefficients in the formula represent the effects of each variable
on the logarithm of the extinction time while keeping the other
variables constant.

For example, an increase in the number of nesting pairs by one unit is
associated with an increase in the logarithm of extinction time by
0.26509 units, holding size and migratory status constant. Similarly,
small-sized birds have a 0.65237 units lower logarithm of extinction
time compared to large-sized birds, holding the number of nesting pairs
and migratory status constant.

This formula can be used as a conclusion to describe the relationship
between the extinction time and the variables of interest (number of
nesting pairs, size, and migratory status) based on the reduced model.
