---
title: "Factors Affecting Extinction"
output: pdf_document
date: "2023-03-04"
---

## Exercise 1
Begin with describing and fitting a full model in which the intercepts and
slopes of the extinction times versus numbers of pairs may be different in all
four combinations of size and migratory status.

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

data <- read.csv("data.csv")

# Create the scatterplot
my_plot <- ggplot(data, aes(x=Pairs, y=Time)) +
  geom_point() +
  facet_wrap(~ Size + Status, nrow=2) +
  geom_smooth(method="lm", se=FALSE)

# Add axis labels and a title
my_plot + xlab("Number of Pairs") + ylab("Years Until Extinction") +
  ggtitle("Relationship between Pairs and Time by Size and Status")
```
Comment: The relationship between years until extinction for a species
and the number of nesting pairs seems to be bigger for migratory species of 
either size than for resident species.

```{r, echo=FALSE}
r_squared <- data %>% 
  group_by(Size, Status) %>% 
  do(model = lm(Time ~ Pairs, data = .)) %>% 
  summarise(
    intercept = coef(model)[1],
    slope = coef(model)[2],
    adj_r_squared = summary(model)$adj.r.squared,
    combination = paste(Size, Status, sep = "_")
  )

# Print R-squared values
print(r_squared)
```

## Exercise 2

```{r}
full_model <- lm(Time ~ Pairs * Size * Status, data = data)

# Create a residual plot
residual_plot <- ggplot(data, aes(x = Pairs, y = resid(full_model))) +
  geom_point() +
  facet_wrap(~ Size + Status, nrow = 2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Number of Pairs") +
  ylab("Residuals") +
  ggtitle("Residual Plot for Full Model")

print(residual_plot)

```
When we examine the residual plot, we should look for the following:

Randomness: The residuals should be randomly scattered around the horizontal dashed line, indicating that there is no pattern to the residuals.

Constant variance: The spread of the residuals should be roughly constant across all values of the predictor variable, indicating that the variance of the response variable is the same for all values of the predictor variable.

Outliers: Points that are far away from the horizontal dashed line may be outliers, indicating that they are not well explained by the model.

Based on the residual plot, we can see that the residuals are mostly randomly scattered around the horizontal dashed line for each group, which is a good sign that the model is fitting the data reasonably well. However, we can also see some patterns in the residuals for some groups, particularly for the small resident and large migrant groups. This may indicate that the model needs to be transformed or that there are outliers affecting the fit of the model for these groups. We should investigate further to determine the cause of these patterns and outliers.




